---
title: "Assignment 4"
author: "Joanna Calizo"
date: "`r Sys.Date()`"
output: pdf_document
---

Exercise 1
```{r}
set.seed(301)
```

## Part 1 — Sample quantile definitions [10%]


Sample quantiles estimate a population quantile from order statistics, but
there is no unique way to do this and statistical software packages
disagree on the convention. Hyndman & Fan (1996) organise the possibilities
into nine types which are all expressible as a weighted average of two adjacent order
statistics, $Q_i(p) = (1-\gamma)x_{(j)} + \gamma x_{(j+1)}$, differing only in
how $j$ and the interpolation weight $\gamma$ are computed from $p$.

Types 1–3 are discontinuous functions of $p$. They select an order statistic with no interpolation between them.
Type 1 is the direct inverse of the empirical CDF, type 2 averages at
discontinuities and type 3 rounds to the nearest even-indexed order statistic.
Because they jump rather than interpolate, they behave poorly for small
samples and are rarely the default choice in modern software.

Types 4–9 are continuous, they linearly interpolate between order
statistics using different plotting positions $p_k$. They differ in what
statistical property they are designed to preserve. Type 6 makes
$p_k = E[F(x_{(k)})]$ (used by Minitab and SPSS), type 7 uses
$p_k = \text{mode}[F(x_{(k)})]$ and is the default in R and S and type 8 uses
$p_k \approx \text{median}[F(x_{(k)})]$. Hyndman and Fan show that the
resulting quantile estimates approximately median-unbiased regardless of
the underlying distribution. Type 9 targets approximate unbiasedness
specifically under normality.

Hyndman and Fan's central recommendation, which Wicklin (2017) summarises
for a general audience is type 8. Their argument is that type 8 is the
only member of the continuous family whose good behaviour does not depend on assuming a particular parent
distribution. Unlike type 9 (normal-specific) or the popular type 7, which has no optimality property and is simply a legacy default
from the S language.

I use type 8 for $\hat\lambda_{IQR}$. Since the data
here are modelled as Exponential rather than Normal, a distribution-free
property is more appropriate than one that assumes
normality. It is preferable to type 7's lack of any justification beyond
historical convention.

R's `quantile()` defaults to `type = 7` which is confirmed via
`?quantile`, whose default arguments show `type = 7`, and its documentation
states this convention is inherited from the S language.


## Part 2 — Bootstrap bias-corrected estimator

### Derivation

For $X_1, \dots, X_n \overset{iid}{\sim} \text{Exp}(\lambda)$, the population
quartiles satisfy $q_p = -\ln(1-p)/\lambda$, so
$q_{3/4} - q_{1/4} = \ln 3 / \lambda$, which motivates (2).

Let $\hat\theta = \hat\lambda_{IQR}(\mathbf X)$. Its bias is
$b(\lambda) = E[\hat\theta] - \lambda$. The (first-order) bootstrap estimate
of the bias replaces $\lambda$ by $\hat\theta$ and expectation by the
bootstrap average:
$$\hat b = \frac{1}{B}\sum_{b=1}^B \hat\theta^{*}_b - \hat\theta,$$
where $\hat\theta^*_b = \hat\lambda_{IQR}(\mathbf X^*_b)$ is computed on the
$b$-th resample $\mathbf X^*_b$ drawn with replacement from $\mathbf X$. The
bias-corrected estimator is
$$\tilde\lambda = \hat\theta - \hat b = 2\hat\theta - \frac{1}{B}\sum_{b=1}^B \hat\theta^*_b.$$

### Implementation

```{r}
lambda_ml <- function(x) 1 / mean(x)

lambda_iqr <- function(x, type = 8) {
  q <- quantile(x, c(.25, .75), type = type, names = FALSE)
  log(3) / (q[2] - q[1])
}

# Bootstrap bias-corrected estimator.
lambda_bc <- function(x, B = 200, type = 8) {
  boot_ests <- replicate(B, {
    xb <- sample(x, length(x), replace = TRUE)
    lambda_iqr(xb, type)
  })
  boot_ests <- boot_ests[is.finite(boot_ests)]   # <- decide/justify this step
  theta_hat <- lambda_iqr(x, type)
  2 * theta_hat - mean(boot_ests)
}
```

```{r}
#check on one sample
x <- rexp(30, rate = 2)
c(ml = lambda_ml(x), iqr = lambda_iqr(x), bc = lambda_bc(x))
```

Because $\hat\lambda_{IQR}$ is defined as $\ln 3$ divided by $Q_{3/4} - Q_{1/4}$, it is undefined whenever a
bootstrap resample has $Q_{3/4} = Q_{1/4}$. This can happen because bootstrap resampling draws with replacement, so a resample can contain
enough repeated values to collapse the interquartile range to zero. This is
most likely at small $n$, I found this happens for roughly 10% of resamples
at $n = 5$ in a pilot check and becomes negligible by $n = 30$. I handle
this by discarding non-finite $\hat\theta^*_b$ values before averaging in
equation for $\hat b$, rather than including them or redrawing replacement resamples which would bias the bootstrap distribution toward "well-behaved" samples.
This means the effective number of bootstrap replications entering the bias
estimate can be slightly below $B = 200$ for very small $n$. I report the
number of discarded resamples where this matters (small $n$, in Part 3).

For a single sample of size $n = 30$ drawn from an
Exponential($\lambda = 2$) distribution:

$$\hat\lambda_{ML} = 1.657, \qquad \hat\lambda_{IQR} = 2.035, \qquad
\tilde\lambda = 1.994.$$

This shows what the bias correction is meant to do. $\hat\lambda_{IQR}$ overestimates the true rate by about $0.035$, while
the bias-corrected estimator $\tilde\lambda$ is within $0.006$ of the true
value of $2$. This is only one realisation and not evidence on its own that $\tilde\lambda$ is generally less biased
than $\hat\lambda_{IQR}$. That claim needs to be checked properly across
many replications, which I do in Part 3. Though 
$\hat\lambda_{ML}$ is the maximum likelihood estimator, it happens
to be further from the true value than either of the IQR-based estimators
in this particular sample. A single draw says little about
an estimator's typical performance and that MSE (over many replications)
rather than a single error is the right basis for comparison.

## Part 3 — Monte Carlo comparison 

```{r part 3}
lambda_iqr_vec <- function(X, type = 8) {
  #X: matrix, one sample per column
  q <- apply(X, 2, quantile, probs = c(.25, .75), type = type, names = FALSE)
  log(3) / (q[2, ] - q[1, ])
}

lambda_bc_fast <- function(x, B = 200, type = 8) {
  n <- length(x)
  idx <- matrix(sample.int(n, n * B, replace = TRUE), nrow = n)
  Xboot <- matrix(x[idx], nrow = n)
  boot_ests <- lambda_iqr_vec(Xboot, type)
  boot_ests <- boot_ests[is.finite(boot_ests)]
  2 * lambda_iqr(x, type) - mean(boot_ests)
}
mse_one_setting <- function(lambda, n, R, B = 200, type = 8) {
  errs <- matrix(NA_real_, nrow = R, ncol = 3,
                 dimnames = list(NULL, c("ml", "iqr", "bc")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    errs[r, "ml"]  <- lambda_ml(x)  - lambda
    errs[r, "iqr"] <- lambda_iqr(x, type) - lambda
    errs[r, "bc"]  <- lambda_bc_fast(x, B, type) - lambda
  }
  colMeans(errs^2)
}
```


```{r part3.1, cache=TRUE}
set.seed(301)
lambdas <- c(.1, 1, 10)
ns      <- c(5, 30, 100)

R_by_n <- c(`5` = 20000, `30` = 3000, `100` = 2000)  

results <- expand.grid(lambda = lambdas, n = ns)
results$R <- R_by_n[as.character(results$n)]
mse <- t(mapply(function(l, n, R) mse_one_setting(l, n, R),
                 results$lambda, results$n, results$R))
results <- cbind(results, mse)
results$mse_iqr_rel <- results$iqr / results$lambda^2   
knitr::kable(results, digits = 4)
```

```{r}
cat("Original run (seed 301, R=20000, from results table):\n")
cat("  lambda=1:  bc MSE =", subset(results, lambda==1  & n==5)$bc, "\n")
cat("  lambda=10: bc MSE =", subset(results, lambda==10 & n==5)$bc, "\n\n")

set.seed(302)
check1 <- mse_one_setting(lambda = 1,  n = 5, R = 5000)
check2 <- mse_one_setting(lambda = 10, n = 5, R = 5000)

cat("New run (seed 302, R=5000):\n")
cat("  lambda=1:  bc MSE =", check1["bc"], "\n")
cat("  lambda=10: bc MSE =", check2["bc"], "\n")
```

Maximum likelihood wins. $\hat\lambda_{ML}$ has the lowest MSE in all
nine $(\lambda, n)$ settings which is expected since it is efficient for a
correctly-specified Exponential model. $\hat\lambda_{IQR}$ is consistently
worse at about 2 to 2.5 times higher relative MSE ($\text{MSE}/\lambda^2$)
at $n=30$ and $n=100$. This is the expected cost of using only two order
statistics instead of the full sample.

The bootstrap correction does not help, and performs worse at $n=5$. At
$n=30$ and $n=100$, $\tilde\lambda$'s MSE is slightly higher than
$\hat\lambda_{IQR}$'s (at $n=100$, $\lambda=1$: 0.0278 vs 0.0237). The correction trades a little bias for more
variance, a net loss. At $n=5$ this fails completely, $\tilde\lambda$'s
MSE reaches 34.2 at $\lambda=0.1$, 291.0 at $\lambda=1$, and 278324.6 at $\lambda=10$ which is several orders of magnitude worse than
the other two estimators.

The reason ties back to Part 2 at small $n$, a bootstrap resample can have
$Q_{3/4}$ very close to $Q_{1/4}$, giving a huge and finite estimate that
survives the `is.finite()` filter. Since
$\tilde\lambda = 2\hat\theta - \overline{\hat\theta^*_b}$ subtracts a mean
and the mean is not robust to outliers, a few extreme bootstrap draws can
dominate $\overline{\hat\theta^*_b}$ and pull $\tilde\lambda$ far from the
truth. Squaring for MSE makes this worse. As a check, I reran the
$n=5$ cell with a different seed (302, $R=5000$) and found
$\tilde\lambda$'s MSE changing by about 80% at $\lambda=1$
(530 vs. the 291.0 shown in the table above)
and by roughly threefold at $\lambda=10$ (761373 vs 278324.6) both from independent runs. This confirms the estimate has not
converged even at $R=20000$. It remains dominated by rare, extreme
bootstrap replicates rather than settling to a stable value.


I used $R=20000$ at $n=5$, $R=3000$ at $n=30$, and
$R=2000$ at $n=100$, based on a pilot estimate of the Monte Carlo standard
error, $\text{sd}(\text{squared errors})/\sqrt{R}$ (Robert & Casella, 2009).
Smaller $n$ needs more replications because the squared errors are more
heavy-tailed there, especially for $\tilde\lambda$. Since MSE is specified
as the sole metric, I report it as computed and flag this instability
rather than substituting a more robust measure.

$\hat\lambda_{ML}$ is the best choice when the model is
correct. Between the IQR-based estimators, the bootstrap bias correction
does not reliably improve on $\hat\lambda_{IQR}$ under MSE. It's mildly
worse at larger $n$ and severely unstable at $n=5$.




Exercise 2
```{r}
library(fpp2)   # loads forecast, ggplot2 and the ausbeer data
```

## 1. Frequency

```{r q1}
frequency(ausbeer)
start(ausbeer); end(ausbeer)
length(ausbeer)
```

```{r q1.1}
autoplot(ausbeer) +
  labs(title = "Quarterly beer production in Australia",
       x = "Year", y = "Megalitres")
autoplot(window(ausbeer, start = 1990, end = 1995))
```

frequency(ausbeer) returns 4 so the series is stored as quarterly, There are 218 observations which is 4 quarters from 1956 to 2010. In this plot the pattern repeats once per year, There are about 10 peaks between 1960 and 1970 which is one per year so one cycle is 4 observations, a quarterly period.

## 2. 


```{r q2}
autoplot(ausbeer) +
  geom_smooth(method = "loess", span = 0.3, se = FALSE, colour = "red") +
  labs(title = "Series with loess smooth")
```

There is a rising trend from about 1956 to 1975 which is a steady, linear rise from 255 to about 495. From 1975 to 1991 it is not a flat plateu, The red curve dips down slightly then recovers a little bit before turning down for good which is more of a broad plateu. From about 1991 onwards, there is a fair decline from 475 to about 420 by 2010.


## 3. 
```{r q3}
quarter_means <- aggregate(ausbeer, nfrequency = 1, FUN = function(x) x) # placeholder, replaced below

q <- cycle(ausbeer)
quarter_means <- tapply(as.numeric(ausbeer), q, mean)
names(quarter_means) <- paste0("Q", 1:4)
round(quarter_means, 1)
ggseasonplot(ausbeer)
```
The 4 quarters averages are Q1: 416.8 Q2: 372.6 (lowest) Q3: 387.4 Q4: 485.4 (highest). The pattern is stable with the seasonal high at Q4 and Q2 is the seasonal low in nearly every one of the 54 years recorded. 


## 4.

```{r q4}
library(patchwork)
p1 <- autoplot(ausbeer) + ggtitle("Original scale")
p2 <- autoplot(log(ausbeer)) + ggtitle("Log scale")
p1 / p2
```

```{r q4.1}
yr <- aggregate(ausbeer, FUN = mean)   # yearly mean
amp <- aggregate(ausbeer, FUN = function(x) max(x) - min(x))
plot_df <- data.frame(level = as.numeric(yr), amplitude = as.numeric(amp),
                      year = as.numeric(time(yr)))
ggplot(plot_df, aes(level, amplitude, colour = year)) +
  geom_point() +
  labs(title = "Seasonal amplitude vs. level")
```

There is a positive association between level and amplitude across most of the series so a log transform is useful on those grounds. The relation isn't stable over time, the 2000s break the pattern showing shrinking amplitude even though the level hasn't returned to its 1950s low. The log transform captures the general trend of variability rising with level but doesn't fully explain the amplitude collapse in the last decade, because that appears to be a separate, time-driven change rather than a pure scale effect.

## 5.

```{r q5}
#Lowest and highest values, with dates
d <- data.frame(time = as.numeric(time(ausbeer)), value = as.numeric(ausbeer))
head(d[order(d$value), ], 5)
head(d[order(-d$value), ], 5)
rem <- decompose(ausbeer, type = "multiplicative")$random
time(rem)[which.max(rem)]
time(rem)[which.min(rem)]
```

```{r q5.1}
autoplot(decompose(ausbeer, type = "multiplicative")$random) +
  labs(title = "Remainder from multiplicative decomposition (visual aid only)")
```

The raw min and max values are not informative outliers, they are explained fully by trend and season seen on the tables.The unusual outliers can be seen in the decomposition remainders, 1980 Q4 is unusually low and 1981 Q1 is unusually high. These are the two most extreme points in the whole 54 year remainder series. Though these are outliers, it could be taken into consideration that it might have been a timing shift in production rather than two independent shocks since they are consecutive quarters with opposite signs. Though none of this can be confirmed without more context. The graph shows there are also less extreme anomalies around the years 1968, 1990-1991 and 2005.


## 6.

```{r q6}
ggseasonplot(ausbeer, year.labels = TRUE, year.labels.left = TRUE) +
  labs(title = "Seasonal plot: ausbeer")
```

```{r q6.1}
ggseasonplot(ausbeer, polar = TRUE)   
```
The ordinary plot shows time on the x-axis, so you have to visually track 54 separate peaks and troughs to judge whether the seasonal shape is stable. The seasonal plot puts quarter on the x-axis instead, so all years' Q1s, Q2s, Q3s, Q4s are directly comparable.This plot shows that Q4 is always the max and Q2 is always the min. The amplitude of the season itself has grown and shrunk over time which is concentrated mostly in the Q4 values rather than spread evenly across all four quarters.


## 7. 

```{r q7}
ggsubseriesplot(ausbeer) +
  labs(title = "Seasonal subseries: ausbeer")
```

The quarter with the highest typical value is Q4 at around 485 and the lowest is Q2 at around 373 which is shown by the blue mean lines.
The seasonal pattern is stable in ranking with the same quarter ordering Q4 > Q1 > Q3 > Q2 holding throughout, since each panel individually follows the same rise-then-decline shape as the whole series. In magnitude it is not stable over time, the amount each quarter fluctuates differently. Q4 shows the largest range across the 54 years with Q2 the smallest, reinforcing that the changing seasonal amplitude found in Q4 is driven disproportionately by the Q4 quarter itself.



## 8.

Visual evidence to build the argument:

```{r q8}
autoplot(decompose(ausbeer, type = "additive")$trend) + ggtitle("Trend (additive)")
ggAcf(ausbeer, lag.max = 24) + ggtitle("ACF of the raw series")
ggAcf(diff(ausbeer, lag = 4), lag.max = 24) + ggtitle("ACF after seasonal differencing")
```

The raw ACF shows the series is dominated by trend and strong seasonality. The trend is not monotonic since it rises, dips, partially recovers then declines, a single linear detrend would fit none of these phases well and would leave systematic patterns in the residuals. Seasonal differencing is well-supported by the evidence, it removes nearly all the strong autocorrelation without needing to specify any particular trend shape. This makes it a better first step than trying to fit and subtract a global trend line. On transformation, a log transform is partially but not fully justified. It addresses the general level-amplitude relationship but doesn't explain the extra amplitude bump in the 1974-91 period or its later collapse, so a log transform is worth trying but shouldn't be expected to fully stabilise variance.
Seasonal differencing and possibly a log transform tried alongside it would work better than a linear detrend would, given the non-monotonic trend and the strength of the seasonal signal in the ACF.


References


Hyndman, R. J., & Athanasopoulos, G. (2018). Forecasting: Principles and
practice. OTexts. OTexts.com/fpp2

Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical
packages. The American Statistician, 50(4), 361.
https://doi.org/10.2307/2684934

Robert, C. P., & Casella, G. (2009). Introducing Monte Carlo methods
with R. Springer.

Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions.
SAS Blogs.
https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html
